GPU-Beschleunigung der Material Point Method

### 2019-01-26 mscthesis.tex
rest checked + abstract

\newenvironment{code}{\captionsetup{type=algorithm}}{}
\begin{document}
\author{Fabian Meyer}
\title{GPU-Beschleunigung der Material Point Method}
\zweitgutachter{Bastian Krayer, M.Sc.}
\zweitgutachterInfo{(Institut f{\"u}r Computervisualistik, AG Computergraphik)}
% Umschalten der Sprache (fuer englische Rubrikbezeichnungen etc.)

@@

\clearpage
\selectlanguage{english}
\subsubsection*{Abstract}

\noindent Physic

%\vfill~
%ggf. noch'n \vfill~

\clearpage
\tableofcontents

@@

Therefore they propose to use the Fixed Corotated Hyperelasticity energy density:
\begin{equation}
  \Psi_{FCH} = \mu \|\boldsymbol{\Sigma}_E - \boldsymbol{I}\|^2_F + \frac{\lambda}{2}(J_E-1)^2.
\end{equation}
This leads to the primary contour $J=1$ which does not cross the inverted region as only one singular value may be negative.
\begin{flushright}\cite{MPM:INVERT}\end{flushright}
\subsection{Governing equations}
fill with nonsense
deriving the weak form
\subsubsection{Conservation of mass}
Let the Eulerian mass density be $_t\rho(_t\boldsymbol{x},t)$ and its (Lagrangian) pull back be $_0\rho(_0\boldsymbol{x},t)$.
After \ref{eq:volume_integral} they are related as:
\begin{equation} \label{eq:density_pull_back}
  \int _ {_tB _ { \epsilon }}_t\rho(_t\boldsymbol{x},t)d_t\boldsymbol{x} =   \int _ {_0B _ { \epsilon }} {^t_0J}_0\rho(_0\boldsymbol{x},t) d_0\boldsymbol{x}.
\end{equation}
An open ball $ _ { 0 }B _ { \epsilon } $ in the reference configuration will have the same mass as its respective open ball $B_{\epsilon}^t$ in the current configuration. Keep in mind that both refer to a deformed state.

@@

\subsection{General-purpose computing on graphics processing units(GPGPU)}

@@

Therefore, one should prefer register memory for intra-thread caching. Shared-memory should be used for extra caching memory(to reduce register spilling) or inter-thread communication within a block. Global memory should be mostly used at the start and end of a shader.

\textbf{Host communication:} A common performance bottleneck is stalling the GPU with CPU-GPU transfers. The attainable bandwidth for PCIe x16 3.0 between host and device is roughly 12GB/s. While the communication between device and GPU is 224GB/s (GTX 970). This includes doing calculations on the GPU that may run faster on the CPU. The API may also stall if the CPU is not issuing instructions fast enough to the GPU. This may be due to calculations on the CPU or, even worse, calculations depending on GPU results. This thesis tries to avoid PCI-transfers wherever possible.

\begin{table}
  \begin{tabular}{ | l | l | l | l | l |}    \hline
    Memory       &Location on/off chip   &Cached   &Scope                &Lifetime   \\\hline
    Register     &On                     &n/a      &1 thread(no shuffle) &thread     \\\hline
    Local        &Off                    &L2       &1 thread             &thread     \\\hline
    Shared       &On                     &n/a      &block                &block      \\\hline
    Global       &Off                    &L2       &all threads+host     &host alloc \\
    \hline
  \end{tabular}
\caption{GPU Memory \cite{NVIDIA:BEST:PRACTICE}}
\end{table}
\textbf{Memory coalescing:} Between a single thread and a block of threads another important entity exists. NVIDIA GPUs group 32 threads into a synchronous warp, respectively AMD groups 64 threads as a wavefront. Generally they are called subgroups. All operations are issued per subgroup. A memory access uses the L2-caching behavior. A L2-cache-line is 32 bytes long. Thus eight (u)int/floats, two single-precision vectors or a single double-precision vector fit into one cache-line. If all threads of a warp access memory within this cache-line, only one memory request of 32 bytes needs to be made. In contrast, if a warp accesses N different cache-lines, N*32 bytes need to be loaded.

The unified L1/texture-cache acts as a coalescing buffer for the thread of a subgroup (Maxwell architecture \cite{NVIDIA:MAXWELL}). It does not matter in which order the cache-line is accessed by the warp. I.e. random or reversed access on a single cache-line are also coalesced. On a global level the easiest to control access pattern thus becomes consecutive or reversed memory accesses. L2 is a write-back cache. Thus this assessment also qualifies for global memory writes. \cite{NVIDIA:MEMORY_BAND}

\textbf{Shared Memory bank conflicts:} Shared memory is split into 32 banks of 4 bytes. The hardware groups together unique memory accesses on these banks. E.g., if two threads within a subgroup access bank 17, a bank conflict occurs, and the access needs to be split into two groups. The one exception is a broadcast: all threads within a subgroup access the same bank. The number of banks is equal to the warp size on the GTX970 s.t. throughput is maximized if every thread does a unique bank access. \cite{NVIDIA:BEST:PRACTICE}

\textbf{Register spilling:} One should avoid local memory. Local memory is created due to register spilling and stored off chip like global memory, though it is L2-cached. The available number of registers per thread is calculable with the CUDA Occupancy calculator \cite{NVIDIA:OCCUPANCY}. If the occupancy falls off to 0\%, register spilling occurs. The occupancy is the amount of active warps divided by the possible maximum amount of warps on a Streaming Multiprocessor. It allows the SM to schedule between warps: If the current warp can not continue (memory latency, barrier etc.) it will switch to another. This is called a warp stall, table \ref{table:metric}. Occupancy is mostly affected due to resource limits on an SM (shared memory, register) as can be seen in the occupancy calculator. Non optimal occupancy might not have any performance impact at all. \cite{NVIDIA:ACHIEVED_OCCUPANCY}

There is no way to find out the exact number of registers an OpenGL shader uses. The program object assembly output of \textit{nvemulate} \cite{NVIDIA:EMULATE} may give an indicator. It will output an intermediate language of one the families of \textit{NV\_gpu\_programX} (where $X$ stands for the version). TEMP variables are registers. Since \textit{NV\_gpu\_programX} is a family of intermediate languages, there is however no guarentees these are not further optimized. Otherwise close monitoring of the occupancy is recommended. All other metrics discussed are readily available for OpenGL.

\textbf{Block size:} In general the block size should be chosen as small as possible to not hit resource limits. One should start however with a block size of at least 64 as can be seen in the Occupancy calculator. Increasing the block size is useful in combination with shared memory usage. Shared memory accesses are typically faster than global memory accesses. The more threads of a block reuse shared memory data the faster the memory throughput will be. Therefore, operators which involve filtering (also MPM-transfers) can use this behavior.

\textbf{Multiple elements per thread:} Merging threads and simultaneously reducing block size can reduce register pressure, if the algorithm allows it. Otherwise merging sequential and parallel work improves I/O latency hiding. Since the block size is limited by the GPU, some algorithms need to be split in tree structures to be processed. One can not avoid an indirection of writing back and loading again from global memory. Therefore algorithms often improve much further from more sequential work to avoid this indirection.

\textbf{Loop unrolling:} Foremost, loop unrolling reduces instructions and can thus speed up compute-bound work. Moreover, loop unrolling might help the compiler reorder instructions and thus execute code faster. Instructions benefiting greatly from this are global loads. \cite{NVIDIA:PEAK_PERFORMANCE} shows at the example of motion blur and ray-marching that this greatly can increase throughput. Whereas the motion blur example uses a fixed amount of samples and therefore is easily unrollable, ray-marching is a dynamic process. Dynamic loops can however be partially unrolled. In the case of ray-marching this would for instance fetch two samples at once although it might process only one. In short, this merges two consecutive loop iterations. The latency of the second instruction might be partly hidden by the first one and thus speed up the work load. Mind however that loop unrolling can increase register pressure. More registers need to stored and process the relevant code where a dynamic loop reuses the same registers for any iteration.

\begin{flushright}\cite{AMD:GPU_OPEN}\cite{NVIDIA:BEST:PRACTICE}\cite{NVIDIA:PEAK_PERFORMANCE}\end{flushright}
\subsubsection{GPU Metrics}\label{sec:metrics}
The GPU metric topping all others is the raw computing time spend on a shader: $\Delta t_c$. The following factors matter to set $\Delta t_c$ into perspective:
\begin{enumerate}
  \item For the GPU most interesting is the scalability with increasing input data sizes.
  \item \label{it:speedup}Numerical algorithms may spend additional time computing (implicit) quantities to allow for larger physical time steps: $\Delta t_p$. Therefore one might be interested how long the GPU needs to calculate one physical second:
    \begin{equation}
    \Delta t^c_{1s} = \frac{\Delta t_c}{\Delta t_p}.
    \end{equation}
    Otherwise one would calculate the speedup of procedure 1 over procedure 2 for a direct comparision:

    \begin{equation}
      \text{Speedup of Procedure 1} = \frac{\Delta t_{c1}}{\Delta t_{p1}} \frac{\Delta t_{p2}}{\Delta t_{c2}}.
    \end{equation}
\end{enumerate}
\begin{table}
  \begin{tabular}{ | l | l | l |}    \hline
    \textbf{Metric}   & \textbf{Max}(GTX 970) & \textbf{Description} \\ \hline
    Speedup           & -                     & item \ref{it:speedup}\\\hline
    $\Delta t_c$      & -                     & computing time\\\hline
    VRAM SOL\%        & 100                   & memory througput(, w.r.t. to hw.-limit)\\\hline
    SM SOL\%          & 100                   & instruction throughput\\\hline
    L2 SOL\%          & 100                   & L2-cache throughput\\\hline
    Tex SOL\%         & 100                   & L1-cache throughput\\\hline
    L2-Hitrate\%      & 100                   & L2 hit-miss ratio\\\hline
    SM Active\%       & 100                   & at least one warp active avg. over SMs\\\hline
    SM Issue Util.\%  & 100                   & amount of cycles an instr. was issued\\\hline
    SM Occupancy      & 64.0                  & active warps avg. over SMs\\\hline
    Read Active\%     & 100                   & \% of time spend on VRAM reads\\\hline
    Write Active\%    & 100                   & \% of time spend on VRAM writes\\\hline
    \multicolumn{3}{|l|}{\textbf{Warp Stall Reasons:}}\\\hline
    Long Scoreboard\% & 100                   & VRAM latency (mostly reads)\\\hline
    Short Scoreboard\%& 100                   & L2/Shared Mem. latency\\\hline
    Barrier\%         & 100                   & block synchronization \\\hline
    No Instruction\%  & 100                   & instruction fetching or\\
		      &                       & waiting on instruction cache miss\\\hline
  \end{tabular}
\caption{GPU Performance Metrics}\label{table:metric}
\end{table}
The other GPU metrics are acquired using NVIDIA Nsight with OpenGL Performance Markers \cite{NVIDIA:METRICS}. An active warp is also one that waits for memory.

None of these metrics necessarily have a correlation to computing time. However they are good indicators on which operations/hardware units the GPU spends its time on. The top SOL\% metric measures fraction of the throughput to the bandwidth of that hardware unit. NVIDIA aims to get the top SOL\% metric between 60\% < top SOL\% < 80\%. Occupancy was discussed in \ref{sec:optim}. \cite{NVIDIA:PEAK_PERFORMANCE}

Aiming to get shaders memory-bandwidth limited while minimizing the amount of needed accesses for any given workload is the top-priority for simple tasks. Achieving memory-bandwidth limit is equal to bringing the VRAM SOL\% between 60\% < top SOL\% < 80\%. The metrics Read Active\%, Write Active\% and Long Scoreboard\% will simultaneously increase by following that aim.

The metric SM Issue Utilization (per active cycle) is important for compute-bound work as memory-bandwidth limited work will spend most of its time outside of the SM fetching or writing data. This metric tells if the SM can issue instructions fast. NVIDIA also aims between 60\% to 80\% for this metric if the shader is compute-bound. Furthermore for compute-bound work often the top warp stall reason is waiting on instruction fetches.

@@

Coalescing leads to the motivation of using structure of arrays(SoA) over arrays of structures(AoS) memory layouts. The latter is the more commonly taught approach, while high performance CPU code may also prefer SoA or the more advanced entity component systems. This opens up the field of data-oriented design which is too big too cover and not main part of this thesis.

In short one might not access every variable of a particle in a shader. I.e. in physics one has a pipeline of different passes each requiring another subset of variables. Assume for instance one shader only requires the position then it is more favorable for consecutive threads to load consecutive position vectors. The cache-line is then filled with position vectors, which all will get used. In contrast in an AoS layout the cache-line will have a position and a velocity loaded, but the shader never uses the velocity, halfing the throughput. This process is show in Figure \ref{fig:soa} and \ref{fig:aos}.

@@

Furthermore we experienced that for sizes starting with 256 KB the NVIDIA-Compiler could not link the OpenGL program using the SoA Layout as in Listing \ref{llst:soa}. As an alternative we had to provide meta information in the style of reflection macros and collapse the container, Listing \ref{llst:opengl}. We can then access SoA by memory offset and AoS by variable or by memory offset for a more abstract view of the memory. We also share the AoS setup between C++ and OpenGL code. One could also build a shader using the abstract layout to convert from SoA to AoS.

@@

    \texttt{\#define binary\_op(x,y) x+y}
\end{enumerate}
For debugging we output the shader with all processed includes and let the gcc-preprocessor assemble the shader with options: \texttt{gcc -E -x c -P}. Except for the omission of the glsl version or extension macros, we did not find any problems in this process.

@@

\textbf{Parallel Reduction}: Assuming an associative \texttt{binary\_op(x,y)}:= $x \circ y$ and an array of values $\left[a_0,a_1, ... ,a_n\right]$, parallel reduction computes:
\begin{equation}
  r = a_0 \circ a_1 \circ ... \circ a_n.
\end{equation}
Our OpenGL parallel reduction implementation uses \cite{NVIDIA:PARALLEL_REDUCTION} as a reference. To avoid memory padding in shared memory the addressing scheme is given by the sequential addressing scheme in Figure \ref{fig:sequential}.
\begin{figure}[htbp]
    \centering
  \includesvg[svgpath =images/,width=\textwidth]{sequential}
  \caption{Sequential addressing for \texttt{vec4}s. Even threads are blue, odd are yellow. Shared-memory banks per instruction are colored: The banks of the first instruction are green and red numbers. The second instruction is only green. The first and all following instructions can be handled fully concurrently. No memory padding required.}
\label{fig:sequential}
\end{figure}

Parallel reduction schemes using shuffle-operations require interleaved addressing due to memory locality restrictions, Figure \ref{fig:interleaved}. They use registers instead of shared memory and thus do not need memory padding. \cite{NVIDIA:SHUFFLE}

\cite{NVIDIA:PARALLEL_REDUCTION} also makes heavy use of combining parallel and sequential work. Every thread adds a multiple of two sequential adds into shared memory. After this the thread does its parallel work. Therefore the reduction factor one invocation does is multiple times higher and allows higher input data sizes without losing any performance, refer to chapter \ref{sec:eval}.

Parallel reduction solves assuming a \texttt{unary\_op(x)} directly after memory load:
\begin{itemize}
  \item CFL-condition of chapter \ref{sec:cfl} with operators
  \begin{itemize}
  \item \texttt{unary\_op(x) := length(x)} and
  \item \texttt{binary\_op(x,y) := max(x,y)}.
  \end{itemize}
\item Discretized conservation of governing equations between time steps or transfers in chapter \ref{sec:mass} and \ref{sec:apic} with
\begin{itemize}
  \item \texttt{unary\_op(x) := x} and
  \item \texttt{binary\_op(x,y) := x + y}.
  \end{itemize}
\item Calculate the maximum count of particles of all grid nodes within a block (\textit{per-block parallel reduction} \ref{sec:active_blocks}):
\begin{itemize}
  \item \texttt{unary\_op(x) := x} and
  \item \texttt{binary\_op(x,y) := max(x,y)}.
\end{itemize}
\end{itemize}

>>>>

\newenvironment{code}{\captionsetup{type=algorithm}}{}

\setcounter{biburllcpenalty}{7000}
\setcounter{biburlucpenalty}{8000}

\begin{document}
\author{Fabian Meyer}
\title{GPU-Beschleunigung der Material Point Method}
\zweitgutachter{Bastian Krayer, M.Sc.}
\zweitgutachterInfo{(Institut f{\"u}r Computervisualistik, AG Computergraphik)}
% Umschalten der Sprache (fuer englische Rubrikbezeichnungen etc.)

@@

\clearpage
\selectlanguage{english}

\vspace*{\fill}~
\begin{center}
\subsubsection*{Abstract}
\end{center}
The Material-Point-Method is a technique allowing for physically accurate simulations. It has found its way into computer graphics and since then rapidly expanded. The Material Point Method's hybrid use of Lagrangian particles as a persistent storage and a background uniform Eulerian grid enable solving of various partial differential equations with ease. It has found its way into computer graphics and since then rapidly expanded.

\noindent Core of this process are the grid and particles transfers that interpolate between the two structures.  These transfers are executed multiple times per frame so preprocessing steps might be taken if they are outweighed.

\noindent The Method however suffers from high execution times and make the method only viable for hero shots. The Material-Point-Method including the transfers are highly parallelizable. As such this thesis proposes how to accelerate the Material-Point-Method using GPGPU techniques.

\noindent Deep sorting with counting sort increases coalescing and L2 hit rates. Binning allows for shared memory filtering techniques using blocking. As another preprocessing step only grid blocks are executed which have particles in them. All transfers do not rely on fixed size bins.
\vfill
\clearpage
\tableofcontents

@@

Therefore they propose to use the Fixed Corotated Hyperelasticity energy density:
\begin{equation}
  \Psi_{FCH} = \mu \|\boldsymbol{\Sigma}_E - \boldsymbol{I}\|^2_F + \frac{\lambda}{2}(J_E-1)^2.
\end{equation}
This leads to the primary contour $J=1$ which does not cross the inverted region as only one singular value may be negative.
\begin{flushright}\cite{MPM:INVERT}\end{flushright}
\subsection{Governing equations}
fill with nonsense
deriving the weak form
\subsubsection{Conservation of mass}
Let the Eulerian mass density be $_t\rho(_t\boldsymbol{x},t)$. Similarly, let its (Lagrangian) pull back be $_0\rho(_0\boldsymbol{x},t)$.
After \ref{eq:volume_integral} they are related as:
\begin{equation} \label{eq:density_pull_back}
  \int _ {_tB _ { \epsilon }}_t\rho(_t\boldsymbol{x},t)d_t\boldsymbol{x} =   \int _ {_0B _ { \epsilon }} {^t_0J}_0\rho(_0\boldsymbol{x},t) d_0\boldsymbol{x}.
\end{equation}
An open ball $ _ { 0 }B _ { \epsilon } $ in the reference configuration will have the same mass as its respective open ball $B_{\epsilon}^t$ in the current configuration. Keep in mind that both refer to a deformed state.

@@

\subsection{General-purpose computing on graphics processing units}

@@

Therefore, one should prefer register memory for intra-thread caching. Shared-memory should be used for extra caching memory(to reduce register spilling) or inter-thread communication within a block. Global memory is the persistent storage on the GPU.

\textbf{Host communication:} A common performance bottleneck is stalling the GPU with CPU-GPU transfers. The attainable bandwidth for PCIe x16 3.0 between host and device is roughly 12GB/s while the communication between device and GPU is 224GB/s (GTX 970). This includes doing calculations on the GPU that may run faster on the CPU. The API may also stall the GPU if the CPU is not issuing instructions fast enough to the GPU. This may be due to calculations on the CPU or, even worse, calculations depending on GPU results. This thesis tries to avoid PCI-transfers wherever possible.

\begin{table}
  \begin{tabular}{ | l | l | l | l | l |}    \hline
    Memory       &Location on/off chip   &Cached   &Scope                &Lifetime   \\\hline
    Register     &On                     &n/a      &1 thread(no shuffle) &thread     \\\hline
    Local        &Off                    &L2       &1 thread             &thread     \\\hline
    Shared       &On                     &n/a      &block                &block      \\\hline
    Global       &Off                    &L2       &all threads+host     &host alloc \\
    \hline
  \end{tabular}
\caption{GPU Memory \cite{NVIDIA:BEST:PRACTICE}}
\end{table}
\textbf{Memory coalescing:} Between a single thread and a block of threads another important entity exists. NVIDIA GPUs group 32 threads into a synchronous warp, respectively AMD groups 64 threads as a wavefront. In general they are called subgroups. All operations are issued per subgroup. A memory access uses the L2-caching behavior. A L2-cache-line is 32 bytes long. Thus eight (u)int/floats, two single-precision vectors or a single double-precision vector fit into one cache-line. If all threads of a warp access memory within this cache-line, only one memory request of 32 bytes needs to be made. In contrast, if a warp accesses N different cache-lines, N*32 bytes need to be loaded.

The unified L1/texture-cache acts as a coalescing buffer for the thread of a subgroup (Maxwell architecture \cite{NVIDIA:MAXWELL}). It does not matter in which order the cache-line is accessed by the warp. I.e. random or reversed access on a single cache-line are also coalesced. On a global level the easiest to control access pattern thus becomes consecutive or reversed memory accesses. L2 is a write back cache. Thus this assessment also qualifies for global memory writes. \cite{NVIDIA:MEMORY_BAND}

\textbf{Shared Memory bank conflicts:} Shared memory is split into 32 banks of 4 bytes. The hardware groups together unique memory accesses on these banks. E.g., if two threads within a subgroup access bank 17, a bank conflict occurs, and the access needs to be split into two groups. The one exception is a broadcast: All threads within a subgroup access the same bank. The number of banks is equal to the warp size on the GTX970 s.t. throughput for 32 floats is maximized if every thread does a unique bank access. \cite{NVIDIA:BEST:PRACTICE}

\textbf{Register spilling:} Local memory should be avoided. Local memory is created due to register spilling and stored off chip like global memory, though it is L2-cached. The available number of registers per thread is calculable with the CUDA Occupancy Calculator \cite{NVIDIA:OCCUPANCY} on NVIDIA hardware. If the occupancy falls off to 0\%, register spilling occurs. The occupancy is either given absolute due the amount of active warp or relative divided by the possible maximum amount of warps on a Streaming Multiprocessor. Occupancy allows the SM to schedule between warps: If the current warp can not continue (memory latency, barrier etc.), the warp scheduler will switch to another. This is called a warp stall, table \ref{table:metric}. Occupancy is mostly affected due to resource limits on Streaming Multiprocessors (shared memory, register), as can be seen in the Occupancy Calculator. Non optimal occupancy might not have any performance impact at all when the GPU has enough warps to switch to at all times. \cite{NVIDIA:ACHIEVED_OCCUPANCY}

There is no exact way to find out the exact number of registers for any OpenGL shader. The program object assembly output of \textit{nvemulate} \cite{NVIDIA:EMULATE} may give an indicator. It will output an intermediate language of one of the families of \textit{NV\_gpu\_\-programX} (where $X$ stands for the version). TEMP variables are registers. Since \textit{NV\_gpu\_programX} is a family of intermediate languages, there is however no guarentees these are not further optimized. Otherwise, close monitoring of the occupancy is recommended. All other metrics discussed are readily available for OpenGL.

\textbf{Block size:} In general the block size should be chosen as small as possible to not hit resource limits. One should start however with a block size of at least 64, as can be seen in the Occupancy Calculator. Increasing the block size is useful in combination with shared memory usage. Shared memory accesses are typically faster than global memory accesses. The more threads of a block reuse shared memory data, the faster the memory throughput will be. Therefore, operators which involve filtering (also MPM-transfers) can use this behavior.

\textbf{Multiple elements per thread:} Merging threads and simultaneously reducing block size can reduce register pressure if the algorithm allows it. Otherwise, merging sequential and parallel work improves I/O latency hiding. Since the block size is limited by the GPU, some algorithms need to be split in tree structures to be processed. An indirection cannot be avoided for writing back and loading again from global memory. Therefore, algorithms often improve much further from more sequential work to avoid this indirection.

\textbf{Loop unrolling:} Foremost, loop unrolling reduces instructions and can thus speed up compute-bound work. Moreover, loop unrolling might help the compiler reorder instructions and thus execute code faster. Instructions benefiting greatly from this are global loads. \cite{NVIDIA:PEAK_PERFORMANCE} shows at the example of motion blur and ray-marching that this greatly can increase throughput. Whereas the motion blur example uses a fixed amount of samples and therefore is easily unrollable, ray-marching is a dynamic process. Dynamic loops can however be partially unrolled. In the case of ray-marching this would for instance fetch two samples at once although it might process only one. In short, this merges two consecutive loop iterations. The latency of the second instruction might be partly hidden by the first one and thus speed up the work load. Mind however, that loop unrolling can increase register pressure. More registers need to be stored and process the relevant code where a dynamic loop reuses the same registers for any iteration.

\begin{flushright}\cite{AMD:GPU_OPEN}\cite{NVIDIA:BEST:PRACTICE}\cite{NVIDIA:PEAK_PERFORMANCE}\end{flushright}
\subsubsection{GPU Metrics}\label{sec:metrics}
The GPU metric topping all others is the raw computing time spend on a shader: $\Delta t_c$. The following factors matter to set $\Delta t_c$ into perspective:
\begin{enumerate}
  \item For the GPU most interesting is the scalability with increasing input data sizes. All algorithms here however scale linear with the input size.
  \item \label{it:speedup}Numerical algorithms may spend additional time computing (implicit) quantities to allow for larger physical time steps: $\Delta t_p$. Therefore one might be interested how long the GPU needs to calculate one physical second:
    \begin{equation}
    \Delta t^c_{1s} = \frac{\Delta t_c}{\Delta t_p}.
    \end{equation}
    Otherwise, one would calculate the speedup of procedure 1 over procedure 2 for a direct comparison:

    \begin{equation}
      \text{Speedup of Procedure 1} = \frac{\Delta t_{c1}}{\Delta t_{p1}} \frac{\Delta t_{p2}}{\Delta t_{c2}}.
    \end{equation}
\end{enumerate}
\begin{table}
  \begin{tabular}{ | l | l | l |}    \hline
    \textbf{Metric}   & \textbf{Max}(GTX 970) & \textbf{Description} \\ \hline
    Speedup           & -                     & item \ref{it:speedup}\\\hline
    $\Delta t_c$      & -                     & computing time\\\hline
    VRAM SOL\%        & 100                   & memory througput w.r.t. to hw.-limit\\\hline
    SM SOL\%          & 100                   & instruction throughput\\\hline
    L2 SOL\%          & 100                   & L2-cache throughput\\\hline
    Tex SOL\%         & 100                   & L1-cache throughput\\\hline
    L2-Hitrate\%      & 100                   & L2 hit-miss ratio\\\hline
    SM Active\%       & 100                   & one warp active avg. over SMs\\\hline
    SM Issue Util.\%  & 100                   & amount of cycles an instr. was issued\\\hline
    SM Occupancy      & 64.0                  & active warps avg. over SMs\\\hline
    Read Active\%     & 100                   & \% of time spend on VRAM reads\\\hline
    Write Active\%    & 100                   & \% of time spend on VRAM writes\\\hline
    \multicolumn{3}{|l|}{\textbf{Warp Stall Reasons:}}\\\hline
    Long Scoreboard\% & 100                   & VRAM latency (mostly reads)\\\hline
    Short Scoreboard\%& 100                   & L2/Shared Mem. latency\\\hline
    Barrier\%         & 100                   & block synchronization \\\hline
    No Instruction\%  & 100                   & instruction fetching or\\
		      &                       & waiting on instruction cache miss\\\hline
  \end{tabular}
\caption{GPU Performance Metrics}\label{table:metric}
\end{table}
The other GPU metrics in table \ref{table:metric} are acquired using NVIDIA Nsight with OpenGL Performance Markers \cite{NVIDIA:METRICS}.

None of these metrics necessarily have a correlation to computing time. They are however good indicators on which operations/hardware units the GPU spends its time on. The top SOL\% metric measures fraction of the throughput to the bandwidth of that hardware unit. NVIDIA aims to get the top SOL\% metric between 60\% and 80\%. Occupancy was discussed in \ref{sec:optim}. \cite{NVIDIA:PEAK_PERFORMANCE}

Aiming to get shaders memory-bandwidth limited, and minimizing the amount of needed accesses for any given workload, is the top priority for simple tasks. Achieving memory-bandwidth limit is equal to bringing the VRAM SOL\% between 60\% and 80\%. The metrics Read Active\%, Write Active\% and Long Scoreboard\% will simultaneously increase by following that aim.

The metric SM Issue Utilization (per active cycle) is important for compute-bound work, as memory-bandwidth limited work will spend most of its time outside of the SM fetching or writing data. This metric tells if the SM can issue instructions fast. NVIDIA also aims between 60\% to 80\% for this metric if the shader is compute-bound. Furthermore, for compute-bound work often the top warp stall reason is waiting on instruction fetches.

@@

Coalescing leads to the motivation of using structure of arrays(SoA) over arrays of structures(AoS) memory layouts. The latter is the more commonly taught approach while high performance CPU code may also prefer SoA or the more advanced entity component systems. This opens up the field of data-oriented design which is too big too cover and not main part of this thesis.

In short, one might not access every variable of a particle in a shader. I.e. in most compute work one has a pipeline of different passes each requiring another subset of the variables. Assume for instance, one shader only requires the position. Then it is more favorable for consecutive threads to load consecutive position vectors. The cache-line is then filled with position vectors which all will get used. In contrast, in an AoS layout the cache-line will have a position and a velocity loaded but the shader never uses the velocity. This halfes the throughput. This process is shown in Figure \ref{fig:soa} and \ref{fig:aos}.

@@

Furthermore, for sizes starting with 256 KB the NVIDIA-Compiler could not link the OpenGL program using the SoA Layout as in Listing \ref{llst:soa}. As an alternative meta information is provided in the style of reflection macros. Therefore, the container is collapsed as in Listing \ref{llst:opengl}. Memory is then accessible by SoA by memory offset and by AoS by variable or by memory offset for a more abstract view of the memory. The AoS setup is shared between C++ and OpenGL code. One could also build a shader using the abstract layout to convert from SoA to AoS. For more coherency converting the whole process to reflection macros is sensible.

@@

    \texttt{\#define binary\_op(x,y) x+y}
\end{enumerate}
For debugging the shader is output with all processed includes. The gcc-pre\-pro\-ces\-sor can preprocess the shader with options: \texttt{gcc -E -x c -P}. Except for the omission of the glsl version or extension macros no problems were encountered in this process.

@@

\textbf{Parallel Reduction}: Assuming an associative \texttt{binary\_op(x,y)}:= $x \circ y$ and an array of values $\left[a_0,a_1, ... ,a_n\right]$ reduction computes:
\begin{equation}
  r = a_0 \circ a_1 \circ ... \circ a_n.
\end{equation}
This OpenGL parallel reduction implementation uses \cite{NVIDIA:PARALLEL_REDUCTION} as a reference. To avoid memory padding in shared memory the addressing scheme is given by the sequential addressing scheme in Figure \ref{fig:sequential}.
\begin{figure}[htbp]
    \centering
  \includesvg[svgpath =images/,width=\textwidth]{sequential}
  \caption{Sequential addressing for \texttt{vec4}s. Even threads are blue. Odd threads are yellow. Shared-memory banks per instruction are colored: The banks of the first instruction are green and red numbers. The second instruction is only green. The first and all following instructions can be handled fully concurrently. No memory padding is required.}
\label{fig:sequential}
\end{figure}

Parallel reduction schemes using warp shuffle-operations require interleaved addressing due to memory locality restrictions, Figure \ref{fig:interleaved}. \cite{NVIDIA:SHUFFLE}

\cite{NVIDIA:PARALLEL_REDUCTION} also make use of combining parallel and sequential work. Every thread adds a multiple of two sequential adds into shared memory. After this the thread does its parallel work. Therefore the reduction factor one invocation does is multiple times higher and allows higher input data sizes without losing any performance, refer to chapter \ref{sec:eval}.

Parallel reduction solves, assuming a \texttt{unary\_op(x)} directly after memory load, the following problems in MPM:
\begin{itemize}
  \item CFL-condition of chapter \ref{sec:cfl} with operators,
  \begin{itemize}
  \item \texttt{unary\_op(x) := length(x)} and
  \item \texttt{binary\_op(x,y) := max(x,y)}.
  \end{itemize}
\item discretized conservation of governing equations between time steps or transfers in chapter \ref{sec:mass} and \ref{sec:apic} with
\begin{itemize}
  \item \texttt{unary\_op(x) := x} and
  \item \texttt{binary\_op(x,y) := x + y}.
  \end{itemize}
\item calculate the maximum count of particles of all bins within a block (\textit{per-block parallel reduction} \ref{sec:active_blocks}):
\begin{itemize}
  \item \texttt{unary\_op(x) := x} and
  \item \texttt{binary\_op(x,y) := max(x,y)}.
\end{itemize}
\end{itemize}

\textbf{Scan:}  Assuming an associative \texttt{binary\_op(x,y)}:= $x \circ y$ , a neutral element $e$ of the \texttt{binary\_op}, and an array of values $\left[a_0,a_1, ... ,a_n\right]$ an exclusive scan computes the array:
\begin{equation}
  \left[e , a_0 , \left(a_0 \circ a_1\right), \left(a_0 \circ a_1 \circ a_2\right), \ ... \ ,\left(a_0\circ a_1 \circ a_2 \circ ...\circ a_{n-1}\right)\right].
\end{equation}
\begin{wrapfigure}[13]{r}{0.20\textwidth}
\vspace{-11}
 \includesvg[svgpath =images/,width=0.20\textwidth]{scan}
 \caption{Scan (add)}
  \label{fig:scan}
\end{wrapfigure}
The scan uses an interleaved addressing pattern, Figure \ref{fig:interleaved}. The interleaved addressing pattern has the benefit over the sequential one of already computing the partial scans for all odd indices (although misaligned). Note, that the reduction needs to reduce to the right achieved by writing into the right element instead of the left in Figure \ref{fig:scan}. A full scan is then computed as \cite{NVIDIA:SCAN}:
\begin{enumerate}
  \item \label{it:up-sweep} Upsweep-Phase: Top-to-Bottom reduction to the right of the array.
  \item Inserting the neutral element into the last element of the array.
  \item Downsweep-Phase: Reverse execution order of item \ref{it:up-sweep} (inverted tree), and in addition overwrite the left element with the right one's.
\end{enumerate}
Again, sequential work can be combined with parallel work. \cite{NVIDIA:SCAN_MODERN} name the process: Raking. The general technique is also called register blocking. Instead of loading one element per thread, one thread loads N elements and computes a sequential partial scan in register memory. The reduced last value of the array is written into shared memory and the parallel scan begins. Afterwards the parallel scan result is spread out(raking) with the sequential partial scan.\cite{NVIDIA:SCAN_MODERN}
\begin{figure}[htbp]
  \includesvg[svgpath =images/,width=\textwidth]{interleaved}
  \caption{Interleaved addressing \texttt{vec4}s. Even threads are blue, odd are yellow. Shared-memory banks per instruction are colored: The banks of the first instruction are green and red numbers. The second instruction is only green. Memory padding needs to be done for concurrent shared-memory access. The first and all following instructions can be handled fully concurrently.}
\label{fig:interleaved}
\end{figure}
As mentioned, the interleaved addressing requires a memory padding for conflict-free addresses. For block sizes of 32 - 512 this becomes:
\begin{equation}\label{eq:bank}
  p = u + (u << \log_2(n)), \text{ or }   p = u + \frac{u}{n}
\end{equation}
where $p$ is the padded memory offset, $u$ is the unpadded memory offset and $n$ is the number of banks.
For a block size of 1024 additional padding is needed for a full conflict-free access:
\begin{equation}\label{eq:bank_1024}
  p = u + (u << \log_2(n)) + (u << (2 * \log_2(n))), \text{ or }   p = u + \frac{u}{n} + \frac{u}{n*n}.
\end{equation}
However, the added computation time does not outweigh the bank-conflicts. So the process is reverted to eq. \ref{eq:bank}. Note, that this is a correction to \cite{NVIDIA:SCAN} where C++-Operator-Precedence of $+$ over $<<$ is not recognized, and additionally incorrectly $ u << n$ is used in the first part of eq. \ref{eq:bank_1024}.
